Natural sources of vagueness and their implications

### 2017-05-11 book_chapter/paper.tex
extension paragraph on rationality

Although frequently debated, especially in areas like philosophy, psychology, and economics, rationality is a notion for which clear definitions are hard to come by.
Discussions around it usually rely on common connotations and implicit assumptions.
The aforementioned classical picture of language and meaning is focused on an aspect of rationality that we can call \emph{procedural}.
It attempts to characterize a particular choice of an agent as rational based on the procedures used to select it.
We can also focus instead on the consequences, rather than the procedures.
\emph{Instrumental} rationality, as it is typically called, characterizes an agent's choice as rational if, in making that choice, the available means were employed in the best way possible to achieve a certain goal.
%This is close to the notion of rationality that is used in economics and game theory: agents are rational if and only if they make decisions that maximize their expected utility.
The notion is linked to David Hume~\parencite*{hume_treatise_1738} and his assertion that ``Reason is, and ought to only be the slave of the passions.''
Another assumption that is important to qualify when talking about rationality of an agent's choice is the amount of information that the agent has access to, when making his decision.
Depending on how accurate and complete the agent's knowledge of its environment, the objectives to be achieved, the choices available, the relation between those choices and the objectives, etc, the verdict over the rationality of his choices can vary.
This is difficult to characterize in general, but, in the context of a given model, we can describe agents with respect to how much knowledge they have of the ingredients of the model.
An \emph{omniscient} agent is one that is in possession of the same information as the modeler, whereas an agent with less than that is said to only have \emph{limited} awareness of the relevant aspects of the model.
Finally, the third assumption we will draw attention to, for now, regards the ability of an agent to take predictions into account.
Thus we can say that an agent is more or less \emph{strategic} depending on the extent to which he is able to anticipate and consider either the actions of other agents, or considerations of medium/long term gains.
The classical picture is, in these terms, tied to a procedural account of rationality, where agents are pretty much omniscient.
Questions of how strategic they are do not really come to the fore, since the analysis is typically both atemporal and individualistic.
\mf{I suggest to rewrite this paragraph, keeping as much as possible, but to include a formal definition of ``instrumental rationality'' as maximization of expected utilities, and to also include notions ``ecological rationality'' and ``bounded rationality'' with special-purpose and clearly spelled out definitions, namely roughly like this: where ``instrumental rationality'' requires a single choice to maximize EU in a single local context / a single stage-game, \emph{ecological rationality} requires a general way of choosing (call it, e.g., a \emph{choice mechanism}) to be optimal (max EU) across a whole class of different contexts / games / decision-making situations; a choice mechanism is \emph{bounded rationality} if it is ecologically rational if we also take into account the processing costs necessary to reach a decision (so a choice mechanism that is fast-and-cheap can be better (under bounded rationality) than a highly precise one on average because it saves processing costs, while it would be strictly outperformed in any game/context/decision situation by a more precise mechanism)}

>>>>

Rationality is an elusive notion which is frequently debated in areas like philosophy, psychology, and economics.
Discussions around it usually rely on common connotations and implicit assumptions.
The aforementioned classical picture of language and meaning is focused on an aspect of rationality that we can call \emph{procedural}.
It attempts to characterize a particular choice of an agent as rational based on the procedures used to select it.
We can also focus instead on the consequences, rather than the procedures.
\emph{Instrumental} rationality, as it is typically called, characterizes an agent's choice as rational if, in making that choice, the available means were employed in the best way possible to achieve a certain goal.
The notion is linked to David Hume~\parencite*{hume_treatise_1738} and his assertion that ``Reason is, and ought to only be the slave of the passions.''
This is close to a notion of rational choice that is used in economics and game theory: agents are rational if and only if they make decisions that maximize their expected utility. 
Concretely, for a (finite) set of acts $A$, a (finite) set of world states $T$ with prior probability $P \in \Delta(T)$ and a utility function $u \colon A \times T \rightarrow \mathbb{R}$, choice of action $a$ is rational only if the expected utility $EU(a) = \sum_{t\in T} P(t) \, u(a,t)$ of $a$ is at least as high as that of any other $a' \in A$.

Choice rationality is not all there is to rationality though.
An agent's mental representation of the decision situation is also subject to considerations of rationality. 
Depending on how accurate and complete the agent's knowledge of its environment, the objectives to be achieved, the choices available, the relation between those choices and the objectives, etc, the verdict over the rationality of his choices can vary.
This is difficult to characterize in general, but, in the context of a given model, we can describe agents with respect to how much knowledge they have of the ingredients of the model.
An \emph{omniscient} agent is one that is in possession of the same information as the modeler, whereas an agent with less than that is said to only have \emph{limited} awareness of the relevant aspects of the model.
In many cases, agents may impossibly have all decision relevant information.
This can happen when an agent has made only a rather small number of observations, e.g., about how a particular word is used.
But even with \emph{partial information}, an agent can be required to have rational beliefs, i.e., to use all of the information available in the best possible way.
This leads back to procedural rationality, logical inference and inductive learning.
It is very hard to say what exactly rational learning is.
It is easier to gesture at what it is not: ignoring observational evidence; failing to draw simple and obvious conclusions; failing to make simple and obvious generalizations etc.
A further aspect of rational belief formation is very important for our present concerns, namely the ability of an agent to make accurate predictions about how other agents behave.
We can say that an agent is more or less \emph{strategically rational} depending on the extent to which he is able to anticipate and consider either the actions and beliefs of other agents, or considerations of medium/long term gains.
Strategic rationality is important for language use, because it is natural to assume that an agent's believes about how an expression is used normally or how it ought to be used properly will depend, at least in part, on beliefs about how others use a said expression.

We draw attention to two additional aspects of rationality which will be helpful in the following.
Where the above definition of instrumental rationality is usually understood as requiring a single choice of act $a$ to maximize the expected utility in a single concrete decision situation, we might also be interested more generally not in specific single local choices, so to speak, but in general global choice mechanisms.
A choice mechanism in thes sense would be a general way of behaving, \emph{e.g.} a function from choice situations or contexts to single choices in each context.
A notion of \emph{ecological rationality} requires a \emph{choice mechanism} to be optimal in the sense of expected utility maximization, across a whole class of different decision situations or contexts, weighed by their probability of frequency of occurrence.
To make this a bit more transparent, consider the following rough formalization.
If a context $c$ from some set $C$ gives us a decision problem with action set $A_c$, states $T_c$, probabilities of states $P_c \in \Delta(T)$ and utility function $u_c$, then a choice mechanism would be a function $CM \colon C \rightarrow \bigcup_{c \in C} A_c$. A choice mechanism $CM$ would be ecologically rational if it maximizes the

The classical picture is, in these terms, tied to a procedural account of rationality, where agents are pretty much omniscient.
Questions of how strategic they are do not really come to the fore, since the analysis is typically both atemporal and individualistic.

\mf{I suggest to rewrite this paragraph, keeping as much as possible, but to include a formal definition of ``instrumental rationality'' as maximization of expected utilities, and to also include notions ``ecological rationality'' and ``bounded rationality'' with special-purpose and clearly spelled out definitions, namely roughly like this: where ``instrumental rationality'' requires a single choice to maximize EU in a single local context / a single stage-game, \emph{ecological rationality} requires a general way of choosing (call it, e.g., a \emph{choice mechanism}) to be optimal (max EU) across a whole class of different contexts / games / decision-making situations; a choice mechanism is \emph{bounded rationality} if it is ecologically rational if we also take into account the processing costs necessary to reach a decision (so a choice mechanism that is fast-and-cheap can be better (under bounded rationality) than a highly precise one on average because it saves processing costs, while it would be strictly outperformed in any game/context/decision situation by a more precise mechanism)}
